### 2016-05-12 text/Chapter03.tex
Context-aware systems edits (Chapter 3)

\paragraph{Chapter structure and contributions}
\begin{itemize}
\item We start with a characterization of contextual properties. The Section~\ref{sec:applications-structure}
  explains what is a \emph{coeffect} and contrasts it with a better known notion of
  \emph{effect}. It explains what is the nature of properties that can be tracked using
  coeffect systems presented in this thesis.

@@

\item The uniform presentation of the systems is the key contribution of this chapter. We distinguish
  between \emph{flat coeffect} systems (Section~\ref{sec:applications-flat}) and \emph{structural coeffect}
  systems (Section~\ref{sec:applications-structural}). The fact that we find a common structure in
  all systems presented here lets us develop unified coeffect calculi in the upcoming three chapters.

@@

\noindent
As mentioned, this chapter may appear as a collection of disconnected
examples\footnote{The different properties captured by monads may appear similarly
disconnected at first!}. But at the end, we will see that they share a common pattern.

@@

In standard $\lambda$-calculus, variable access can be seen as a primitive operation that
accesses the context. The variable access expression introduces a context requirement -- the
expression $x$ is typeable only in a context that contains $x:\tau$ for some type $\tau$.

The standard lambda abstraction (\emph{pure}), shown in Figure~\ref{fig:applications-abs},
splits the free-variable context of an expression into two parts. At runtime, the value of the parameter
has to be provided by the \emph{call site} (dynamic scope) and the remaining values are provided
by the \emph{declaration site} (lexical scope). In the type checking, the splitting is determined
syntactically -- the notation $\lambda x.e$ names the variable whose value comes from the call site.

@@

We look at resources in distributed programming in more details in Section~\ref{sec:applications-flat-distr}.
The important point here is that in flat coeffect systems, contextual requirements are
\emph{split} between the call site and declaration site. Furthermore, there is no syntactic
structure that determines how the requirements are split. In the case of distributed
programming, the resources can be freely associated with either of the two sites.

@@

Here, the variable context $\Gamma$ contains two variables, both of type \ident{Arr}. This means
$\Gamma = x\!:\!\ident{Arr},\, y\!:\!\ident{Arr}$. For simplicity, we treat \kvd{cursor} as a
language primitive. The coeffect annotations will be $(0,0)$ for $x$ and $(-1,1)$ for $y$,
denoting that we access only the current value in $x$, but we need access to both left and right
neighbours in the $y$ array. In order to unify the flat and structural notions, we attach this information
as a \emph{vector} of annotations associated with a \emph{vector} of variable and write:
$\coctx{x\!:\!\ident{Arr},\, y\!:\!\ident{Arr}}{ \aclrd{\alift{ (0,0), (-1,1) }} }$.
The unification is discussed in Chapter~\ref{ch:unified}.

In structural systems, the splitting of context is determined by the name (variable) binding.
For example, consider a function that takes $y$ and contains the above body:
$\lambda y.x[\kvd{cursor}] = y[\kvd{cursor}-1] + y[\kvd{cursor}+1]$. Here, the declaration site
contains $x$ and needs to provide access at least within a range $(0,0)$. The call site provides
a value for $y$, which needs to be accessible at least within $(-1, 1)$. In this way, structural
coeffects remove the non-determinism arising from the splitting of requirements in flat coeffect systems.

@@

The $\lambda$-calculus is asymmetric. It maps a context with \emph{multiple} variables to a
\emph{single} result. An expression with $n$ free variables of types $\tau_i$ can be modelled by a function
$\tau_1 \times \ldots \times \tau_n \rightarrow \tau$ with a product on the left, but a single value
on the right. In both effect systems and coeffect systems, we \emph{write} the annotation as part of
the function arrow. However, in the underlying categorical model, effects are attached to the result
$\tau$, while coeffects are attached to the context $\tau_1 \times \ldots \times \tau_n$.

Structural coeffects have one coeffect annotation per each variable. Thus, the annotation consists
of multiple values -- one belonging to each variable. To distinguish between the overall annotation
and individual (per-variable) annotations, we call the overall coeffect a \emph{vector} consisting of
\emph{scalar} coeffects. This asymmetry also explains why coeffect systems are not trivially dual to
effect systems.

It is useful to clarify how vectors are used in this thesis. Suppose we have a set $\C$ of
\emph{scalars} ranged over by $\cclrd{r},\cclrd{s},\cclrd{t}$. A vector $\aclrd{R}$
over $\C$ is a tuple $\alift{ \cclrd{r_1}, \ldots, \cclrd{r_n} }$ of scalars.
We use bold face letters like $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ for vectors and
normal face $\cclrd{r},\cclrd{s},\cclrd{t}$ for scalars\footnote{For better readability, the thesis
also distinguishes different structures using colours. However ignoring the colour does not introduce
any ambiguity.}. We also say that a \emph{shape} of a vector $\slen{\aclrd{\textbf{r}}}$ (or more generally any container)
determines the set of \emph{positions} in a vector. So, a vector of a shape (length) $n$ has positions
$\{ 1, 2, \ldots, n \}$. We discuss containers and shapes further in Chapter~\ref{ch:unified} and also
discuss how our use relates to containers of Abbott, Altenkirch and Ghani \cite{types-containers}.

Just as in the usual pointwise multiplication of a vector by a scalar, we lift any binary operation on scalars into a
scalar-vector one. For a binary operation on scalars $\circ : \C \times \C \rightarrow \C$, we define
 $\cclrd{s} \circ \aclrd{\textbf{r}} = \alift{ \cclrd{s}\circ\cclrd{r_1}, \ldots, \cclrd{s}\circ\cclrd{r_n}}$.
Relations on scalars can be also lifted to vectors. Given two vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$ of the
same shape with positions $\{ 1, \ldots, n \}$ and a relation $\varpropto\, \subseteq \C \times \C$ we define
$\aclrd{\textbf{r}} \varpropto \aclrd{\textbf{s}} \Leftrightarrow (\cclrd{r_1} \varpropto \cclrd{s_1}) \wedge \ldots \wedge (\cclrd{r_n} \varpropto \cclrd{s_n}) $
Finally, we often concatenate vectors -- for example, when joining two variable contexts.
Given vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$ with (possibly different) shapes $\{ 1, \ldots, n \}$ and
$\{ 1, \ldots, m \}$, the associative operation for concatenation $\times$ is defined as
$\aclrd{\textbf{r}}\times\aclrd{\textbf{s}} = \alift{\cclrd{r_1},\ldots,\cclrd{r_n},\cclrd{s_1},\ldots,\cclrd{s_m}}$.

@@

For three of the examples, we present a type system and a simple semantics. Although the examples
are not new, our novel presentation of the systems (and the fact that they appear side-by-side)
makes it possible to see that they share a common structure. The structure is captured by
a unified \emph{flat coeffect calculus} in Chapter~\ref{ch:flat}.

@@

\paragraph{Implicit parameters.}
Implicit parameters are a special kind of variables that support dynamic scoping.
They make it possible to parameterise a computation (involving a long chain of function calls)
without passing parameters explicitly as additional arguments of all involved functions.

@@

A simple language with implicit parameters has an expression $\ident{?param}$ to read a parameter
and an expression\footnote{Haskell uses $\kvd{let}~\ident{?p} = e_1~\kvd{in}~e_2$, but we use a
different keyword to avoid confusion.} $\kvd{letdyn}~\ident{?param} = e_1~\kvd{in}~e_2$ that sets a
parameter $\ident{?param}$ to the value of $e_1$ and evaluates $e_2$ in a context containing
$\ident{?param}$.

The fact that implicit parameters support both lexical and dynamic scoping becomes interesting
when we consider nested functions. The following function does some pre-processing and then returns a
function that builds a formatted string based on two implicit parameters $\ident{?width}$ and
$\ident{?size}$:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{format} = \lambda \ident{str}~\rightarrow \\[-0.25em]
\quad \kvd{let}~\ident{lines} = \ident{formatLines}~\ident{str}~\ident{?width}~\kvd{in}\\[-0.25em]
\quad (\lambda \ident{rest}~\rightarrow~\ident{append}~
         \ident{lines}~\ident{rest}~\ident{?width}~\ident{?size})
\end{array}
\end{equation*}
%
The body of the outer function accesses the parameter $\ident{?width}$, so it certainly requires a context
$\{ \ident{?width} : \ident{int} \}$. The nested function (returned as a result) uses the parameter
$\ident{?width}$, but in addition also uses $\ident{?size}$. Where should the parameters used by the
nested function come from?

To keep examples in this chapter uniform, we do not use the Haskell notation and instead
write $\tau_1 \xrightarrow{r} \tau_2$ for a function that requires implicit parameters specified by $r$.
In a purely dynamically scoped system, they would have to be defined when the user invokes the nested function.
However, implicit parameters behave as a combination of lexical and dynamic scoping. This means
that the nested function can capture the value of $\ident{?width}$ and require just $\ident{?size}$.
The following shows the two options:
%
\begin{equation}
\tag{\emph{dynamic}}
\ident{string} \xrightarrow{ \{ \ident{?width} : \ident{int} \} }
  (\ident{string} \xrightarrow{ \{ \ident{?width} : \ident{int}, \ident{?size} : \ident{int} \} } \ident{string})
\end{equation}
\vspace{-1em}
\begin{equation}
\tag{\emph{mixed}}
\ident{string} \xrightarrow{ \{ \ident{?width} : \ident{int} \} }
  (\ident{string} \xrightarrow{ \{ \ident{?size} : \ident{int} \} } \ident{string})
\end{equation}
%
This is not a complete list of possible typings, but it demonstrates the options. The (\emph{dynamic})
case requires the parameter \ident{?width} twice (this may be confusing when the caller provides
two different values). In the (\emph{mixed}) case, the nested function captures the \ident{?width}
parameter available from the declaration site. Using the latter typing, the function can be called as follows:\
%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{formatHello} = (~\kvd{letdyn}~\ident{?width}=5~\kvd{in}~\ident{format}~\texttt{"Hello"})\\[-0.25em]
 \kvd{in}\,(~\kvd{letdyn}~\ident{?size} = 10~\kvd{in}~\ident{formatHello}~\texttt{"world"}~)
\end{array}
\end{equation*}
%
For different typings of \ident{format}, different ways of calling it are valid. This illustrates
the point made in Section~\ref{sec:applications-structure-lam} -- flat coeffect systems may
introduce certain non-determinism in the typing. The following section shows how this looks in the
type system for implicit parameters.

@@

Figure~\ref{fig:applications-flat-impl} shows a type system that tracks the set of expression's
implicit parameters. The type system uses judgements of the form $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$
meaning that an expression $e$ has a type $\tau$ in a free-variable context $\Gamma$ with a set
of implicit parameters specified by $\cclrd{r}$. The annotations $\cclrd{r},\cclrd{s},\cclrd{t}$ are
finite partial functions mapping implicit parameter names to types, \ie~$\cclrd{r},\cclrd{s},\cclrd{t} \subseteq
\ident{Names} \mapsto \ident{Types}$. The expressions include \ident{?param} to read implicit
parameter and \kvd{letdyn} to bind an implicit parameter. The types are standard, but functions are
annotated with the set of implicit parameters that must be available on the call site, \ie~
$\tau_1 \xrightarrow{\cclrd{s}} \tau_2$.

Accessing an ordinary variable (\emph{var}) does not require any implicit parameters. The rule that
introduces primitive context requirements is (\emph{param}). Accessing a parameter \ident{?param}
of type $\tau$ requires it to be available in the context. The context may provide more (unused)
implicit parameters thanks to the (\emph{sub}) rule.

When we read the rules from the top to the bottom, application (\emph{app}) and let binding
(\emph{let}) simply union the context requirements of the sub-expressions. However, lambda abstraction
(\emph{abs}) is where the example differs from effect systems. The implicit parameters required by
the body $\cclrd{r} \cup \cclrd{s}$ can be freely split between the declaration site ($\coctx{\Gamma}{\cclrd{r}}$)
and the call site ($\tau_1 \xrightarrow{\cclrd{s}} \tau_2$). Finally, (\emph{letdyn}) defines an implicit
parameter and removes it from the set of requirements.

@@

\begin{figure}

{\small The semantics is defined inductively over the typing derivation:}
\begin{equation}
\tag{\emph{var}}
\sem{\coctx{\Gamma}{\cclrd{r}} \vdash x_i : \tau_i} = \lambda ((x_1, \ldots, x_n), \_) \rightarrow x_i
\end{equation}
\vspace{-1.5em}
\begin{equation}
\tag{\emph{param}}
\sem{\coctx{\Gamma}{\cclrd{r}} \vdash ?p : \sigma} = \lambda (\_, f) \rightarrow f~?p
\end{equation}
\vspace{-1.5em}
\begin{equation}
\tag{\emph{sub}}
\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau} = \lambda (x, f) \rightarrow
  \sem{\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau}~(x, \restr{f}{\cclrd{r'}})
\end{equation}
\vspace{-1.5em}
\begin{equation}
\tag{\emph{abs}}
\hspace{-0.5em}
\begin{array}{l}
  \sem{\coctx{\Gamma}{\cclrd{r}} \vdash \lambda y. e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } = \lambda ((x_1, \ldots, x_n), f) \rightarrow\\[-0.25em]
  \qquad\lambda (y, g) \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\cclrd{r}\cup\cclrd{s}} \vdash e : \tau_2 }~((x_1, \ldots, x_n, y), f \uplus g)
\end{array}
\end{equation}
\vspace{-1.0em}
\begin{equation*}
\tag{\emph{app}}
\hspace{-0.5em}
\begin{array}{l}
  \sem{\coctx{\Gamma}{\cclrd{r}\cup\cclrd{s}\cup\cclrd{t}} \vdash e_1~e_2 : \tau_2 } = \lambda (x, f) \rightarrow \\[-0.25em]
  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 }~(x, \restr{f}{\cclrd{r}})\\[-0.25em]
  \qquad\kvd{in}~g~(\sem{ \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }~(x, \restr{f}{\cclrd{s}} ), \restr{f}{\cclrd{t}})
\end{array}
\end{equation*}
\vspace{-1.0em}
\begin{equation}
\tag{\emph{letdyn}}
\hspace{-0.5em}
\begin{array}{l}
  \sem{\coctx{\Gamma}{\cclrd{r \cup (s \setminus \{ \ident{?p}:\tau_1 \})}} \vdash \kvd{letdyn}~\ident{?p}=e_1~\kvd{in}~e_2 : \tau_2 } =
     \lambda (x, f) \rightarrow \\[-0.25em]
  \qquad\kvd{let}~v=\sem{ \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 }~(x, \restr{f}{\cclrd{r}})\\[-0.25em]
  \qquad\kvd{in}~\sem{ \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_2 }~
     (x, \restr{f}{\cclrd{s} \setminus \cclrd{\{ \ident{?p}:\tau_1 \}}} \uplus \{ \ident{?p} \mapsto v \} )
\end{array}
\end{equation}

\vspace{1em}
{\small Here $\uplus$ and $\restr{f}{\cclrd{r}}$ are auxiliary definitions:}
\begin{equation*}
\begin{array}{rcl}
 \restr{f}{\cclrd{r}} &=& \{ (p,v) \;|\; (p,v) \in f, \;p \in \cclrd{r} \}\\[-0.25em]
 f \uplus g &=& \restr{f}{\, \textit{dom}(f) \setminus \textit{dom}(g)} \cup g
\end{array}
\end{equation*}

\figcaption{Semantics of a language with implicit parameters}
\label{fig:applications-flat-implsem}
\end{figure}

@@

\paragraph{Semantics.}
Implicit parameters can be implemented by passing around a hidden dictionary that provides values
to the implicit parameters. Accessing a parameter then becomes a lookup in the dictionary and
the new \kvd{letdyn} construct extends the dictionary. To elucidate how such hidden dictionaries
are propagated through the program when using lambda abstractions and applications, we present a
simple semantics for implicit parameters. The goal here is not to prove properties of the language,
but simply to provide a better explanation. A detailed semantics in terms of indexed comonads is
shown in Chapter~\ref{ch:flat}.

For simplicity, we assume that all implicit parameters have a type $\sigma$. In that setting,
coeffect annotations $\cclrd{r}$ are just sets of names, \ie~$\cclrd{r}, \cclrd{s}, \cclrd{t} \subseteq \ident{Names}$.
Given an expression $e$ of type $\tau$ that requires free variables $\Gamma$ and implicit parameters
$\cclrd{r}$, the semantics is a function that takes a product of variables from $\Gamma$ together
with a dictionary of implicit parameters and returns $\tau$:

>>>>

This chapter reviews some of the existing context-aware programming abstractions and presents them
in a uniform way. We start with disconnected examples, but at the end, we will see that they
share a common pattern\footnote{The different properties captured by monads may appear similarly
disconnected at first!}.

\paragraph{Chapter structure and contributions}
\begin{itemize}
\item We characterize contextual properties -- the Section~\ref{sec:applications-structure}
  explains what is a \emph{coeffect} and contrasts it with a better known notion of
  \emph{effect}. It explains what is the nature of properties that can be tracked using
  coeffect systems presented in this thesis.

@@

\item The uniform presentation of the systems is the key contribution of this chapter. We distinguish
  between \emph{flat coeffect} systems (Section~\ref{sec:applications-flat}) and \emph{structural coeffect}
  systems (Section~\ref{sec:applications-structural}). This common structure is precisely
  captured by the two \emph{coeffect calculi} in the upcoming chapters.

@@

@@

As discussed in Section~\ref{sec:path-binding}, variable access can be seen as a basic form
of context requirement in $\lambda$-calculus. The expression $x$ is typeable only in a context
that contains $x:\tau$ for some type $\tau$.

In lexically scoped languages, lambda abstraction (\emph{pure}), as shown in
Figure~\ref{fig:applications-abs}, splits the free-variable context of an expression into two parts.
At runtime, the value of the parameter has to be provided by the \emph{call site} (dynamic scope)
and the remaining values are provided by the \emph{declaration site} (lexical scope). In the type
checking, the splitting is determined syntactically. The notation $\lambda x.e$ \emph{names} the
variable whose value comes from the call site.

@@

We look at resources in distributed programming in more details in Section~\ref{sec:applications-flat-distr}.
The important point here is that in flat coeffect systems, contextual requirements are
\emph{split} between the call site and declaration site. Furthermore, there is no syntactic
structure that determines how the requirements are split. As mentioned in Section~\ref{sec:path-binding-amb},
we decouple the definition of semantics from the domain-specific choice that determines how
context requirements are satisfied. We capture the choice in the type and give semantics over a
\emph{typing derivation}. A domain-specific algorithm then chooses the desirable typing -- for
example, by preferring resources available on the client over resources available on the server.

@@

Here, the variable context $\Gamma$ contains two variables, both of type \ident{Arr}. This means
$\Gamma = x\!:\!\ident{Arr},\, y\!:\!\ident{Arr}$. For simplicity, we treat \kvd{cursor} as a
language primitive. The coeffect annotations will be $(0,0)$ for $x$ and $(-1,1)$ for $y$,
denoting that we access only the current value in $x$, but we need access to both left and right
neighbours in the $y$ array. In order to unify the flat and structural notions, we attach this information
as a \emph{vector} of annotations associated with a \emph{vector} of variable and write:
$\coctx{x\!:\!\ident{Arr},\, y\!:\!\ident{Arr}}{ \aclrd{\alift{ (0,0), (-1,1) }} }$.
The unification is outlined in Section~\ref{ch:further-unified}.

In structural systems, the splitting of context is determined by the name (variable) binding.
For example, consider a function that takes $y$ and contains the above body:
$\lambda y.x[\kvd{cursor}] = y[\kvd{cursor}-1] + y[\kvd{cursor}+1]$. Here, the declaration site
contains $x$ and needs to provide access at least within a range $(0,0)$. The call site provides
a value for $y$, which needs to be accessible at least within $(-1, 1)$. In this way, structural
coeffects remove the ambiguity arising from the splitting of requirements in flat coeffect systems.

@@

The $\lambda$-calculus is asymmetric. It maps a context with \emph{multiple} variables to a
\emph{single} result. An expression with $n$ free variables of types $\tau_i$ can be modelled by a function
$\tau_1 \times \ldots \times \tau_n \rightarrow \tau$ with a product on the left, but a single value
on the right. In both effect systems and coeffect systems, we write the annotation as part of
the function arrow. However, in the underlying categorical model, effects are attached to the result
$\tau$, while coeffects are attached to the context $\tau_1 \times \ldots \times \tau_n$.

Structural coeffects have one annotation per each variable. Thus, the annotation consists
of multiple values -- one belonging to each variable. To distinguish between the overall annotation
and individual (per-variable) annotations, we call the overall coeffect a \emph{vector} consisting of
\emph{scalar} coeffects. This asymmetry also explains why coeffects are not trivially dual to
effects.

It is useful to clarify how vectors are used in this thesis. Suppose we have a set $\C$ of
\emph{scalars} ranged over by $\cclrd{r},\cclrd{s},\cclrd{t}$. A vector $\aclrd{R}$
over $\C$ is a tuple $\alift{ \cclrd{r_1}, \ldots, \cclrd{r_n} }$ of scalars.
We use bold face letters like $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ for
vectors and normal face $\cclrd{r},\cclrd{s},\cclrd{t}$ for scalars. We also say that a
\emph{shape} of a vector $\slen{\aclrd{\textbf{r}}}$ (or more generally any container)
determines the set of \emph{positions} in a vector. So, a vector of a shape (length) $n$ has positions
$\{ 1, 2, \ldots, n \}$. We discuss containers and shapes further in Section~\ref{ch:further-unified} and also
discuss how our use relates to containers of Abbott, Altenkirch and Ghani \cite{types-containers}.

Just as in the usual pointwise multiplication of a vector by a scalar, we lift any binary operation on scalars into a
scalar-vector one. For a binary operation on scalars $\circ : \C \times \C \rightarrow \C$, we define
 $\cclrd{s} \circ \aclrd{\textbf{r}} = \alift{ \cclrd{s}\circ\cclrd{r_1}, \ldots, \cclrd{s}\circ\cclrd{r_n}}$.
Relations on scalars can be also lifted to vectors. Given two vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$ of the
same shape with positions $\{ 1, \ldots, n \}$ and a relation $\varpropto\, \subseteq \C \times \C$ we define
$\aclrd{\textbf{r}} \varpropto \aclrd{\textbf{s}} \Leftrightarrow (\cclrd{r_1} \varpropto \cclrd{s_1}) \wedge \ldots \wedge (\cclrd{r_n} \varpropto \cclrd{s_n}) $
Finally, we often concatenate vectors, for example, when joining two variable contexts.
Given vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$ with (possibly different) shapes $\{ 1, \ldots, n \}$ and
$\{ 1, \ldots, m \}$, the associative operation for concatenation $\times$ is defined as
$\aclrd{\textbf{r}}\times\aclrd{\textbf{s}} = \alift{\cclrd{r_1},\ldots,\cclrd{r_n},\cclrd{s_1},\ldots,\cclrd{s_m}}$.

@@

For three of the examples, we present a type system and a simple semantics (given inductively over
the typing derivation). We informally discuss how preferred typing derivation is chosen (to resolve
the inherent ambiguity), but leave details to later chapters. Although the examples
are not new, our novel presentation of the systems (and the fact that they appear side-by-side)
makes it possible to see that they share a common structure. The structure is captured by
a unified \emph{flat coeffect calculus} in Chapter~\ref{ch:flat}.

@@

\paragraph{Implicit parameters.}
As discussed in Section~\ref{sec:path-binding-impl}, implicit parameters are a special kind of
variables that support dynamic scoping. They make it possible to parameterise a computation
(involving a long chain of function calls) without passing parameters explicitly as additional
arguments of all involved functions.

@@

A simple language with support for implicit parameters has an expression $\ident{?param}$ to read a
parameter and an expression\footnote{Haskell uses $\kvd{let}~\ident{?p} = e_1~\kvd{in}~e_2$, but we use a
different keyword to avoid confusion.} $\kvd{letdyn}~\ident{?param} = e_1~\kvd{in}~e_2$ that sets a
parameter $\ident{?param}$ to the value of $e_1$ and evaluates $e_2$ in a context containing
$\ident{?param}$.

The fact that implicit parameters support both lexical and dynamic scoping becomes interesting
when we consider nested functions. The following function does some pre-processing and then returns a
function that builds a formatted string based on two implicit parameters $\ident{?width}$ and
$\ident{?size}$:

\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{format} = \lambda \ident{str}~\rightarrow \\[-0.25em]
\quad \kvd{let}~\ident{lines} = \ident{formatLines}~\ident{str}~\ident{?width}~\kvd{in}\\[-0.25em]
\quad (\lambda \ident{rest}~\rightarrow~\ident{append}~
         \ident{lines}~\ident{rest}~\ident{?width}~\ident{?size})
\end{array}
\end{equation*}
%
The body of the outer function accesses the parameter $\ident{?width}$, so it certainly requires a context
$\{ \ident{?width} \}$. The nested function (returned as a result) uses the parameter
$\ident{?width}$, but in addition also uses $\ident{?size}$. Where should the parameters used by the
nested function come from?

To keep examples in this chapter uniform, we do not use the Haskell notation and instead
write $\tau_1 \xrightarrow{r} \tau_2$ for a function that requires implicit parameters specified by $r$.
We also assume that are implicit parameters are of type $\ident{num}$, so the annotation can be a
simple set of names (rather than mapping from names to types). In a purely dynamically scoped system,
implicit parameters would have to be defined when the user invokes the nested function.
However, implicit parameters behave as a combination of lexical and dynamic scoping. This means
that the nested function can capture the value of $\ident{?width}$ and require just $\ident{?size}$.
The following shows the two options:
%
\begin{equation}
\tag{\emph{dynamic}}
\ident{string} \xrightarrow{ \{ \ident{?width} \} }
  (\ident{string} \xrightarrow{ \{ \ident{?width}, \ident{?size} \} } \ident{string})
\end{equation}
\vspace{-1em}
\begin{equation}
\tag{\emph{mixed}}
\ident{string} \xrightarrow{ \{ \ident{?width} \} }
  (\ident{string} \xrightarrow{ \{ \ident{?size} \} } \ident{string})
\end{equation}
%
This is not a complete list of possible typings, but it demonstrates the options. The (\emph{dynamic})
case requires the parameter \ident{?width} twice (the caller may provide different value, in which
case, the semantics needs to specify which value is preferred). In the (\emph{mixed}) case, the
nested function captures the \ident{?width} parameter available from the declaration site. Using
the latter typing, the function can be called as follows:
%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{formatHello} = (~\kvd{letdyn}~\ident{?width}=5~\kvd{in}~\ident{format}~\texttt{"Hello"})\\[-0.25em]
 \kvd{in}\,(~\kvd{letdyn}~\ident{?size} = 10~\kvd{in}~\ident{formatHello}~\texttt{"world"}~)
\end{array}
\end{equation*}
%
For different typings of \ident{format}, different ways of calling it are valid. This illustrates
the point made in Section~\ref{sec:applications-structure-lam} -- flat coeffect programs have
multiple typing derivations and the semantics depends on the domain-specific choice of preferred
typing. The following section shows how this looks in the type system for implicit parameters.

@@

Figure~\ref{fig:applications-flat-impl} shows a type system that tracks the set of expression's
implicit parameters. The type system uses judgements of the form $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$
meaning that an expression $e$ has a type $\tau$ in a free-variable context $\Gamma$ with a set
of implicit parameters specified by $\cclrd{r}$. The annotations $\cclrd{r},\cclrd{s},\cclrd{t}$ are
finite partial functions mapping implicit parameter names to types, \ie~$\cclrd{r},\cclrd{s},\cclrd{t} \subseteq
\ident{Names}$. The expressions include \ident{?param} to read implicit
parameter and \kvd{letdyn} to bind an implicit parameter. The types are standard, but functions are
annotated with the set of implicit parameters that must be available on the call site, \ie~
$\tau_1 \xrightarrow{\cclrd{s}} \tau_2$.

Accessing an ordinary variable (\emph{var}) does not require any implicit parameters. The rule that
introduces primitive context requirements is (\emph{param}). Accessing a parameter \ident{?param}
requires it to be available in the context. The context may provide more (unused)
implicit parameters thanks to the subcoeffecting rule (\emph{sub}).

When we read the rules from the top to the bottom, application (\emph{app}) and let binding
(\emph{let}) simply union the context requirements of the sub-expressions. However, lambda abstraction
(\emph{abs}) is where the example differs from effect systems. The implicit parameters required by
the body $\cclrd{r} \cup \cclrd{s}$ can be freely split between the declaration site ($\coctx{\Gamma}{\cclrd{r}}$)
and the call site ($\tau_1 \xrightarrow{\cclrd{s}} \tau_2$) and thus an expression may have multiple
valid typing derivations. Finally, (\emph{letdyn}) removes the bound parameter from the set of requirements.

@@

\begin{figure}[t]
\begin{equation*}
\hspace{0em}
\begin{array}{ll}
\hspace{4.5em}\semdef
  {\coctx{\Gamma}{\cclrd{r}} \vdash x_i : \tau_i}
  {\lambda ((x_1, \ldots, x_n), \_) . x_i}
& (\emph{var})
\\[0.0em]
\hspace{3.5em}\semdef
  {\coctx{\Gamma}{\cclrd{r}} \vdash \ident{?p} : \ident{num}}
  {\lambda (\_, f) . f~\ident{?p}}
& (\emph{param})
\\[1.5em]
\hspace{5em}\semdeff
  {\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau}
  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}
  {f}
  {\lambda (x, g) . f~(x, \restr{g}{\cclrd{r'}})}
& (\emph{sub})
\\[1.5em]
\hspace{0em}\emdeff
  {\coctx{\Gamma,y:\tau_1}{\cclrd{r}\cup\cclrd{s}} \vdash e : \tau_2}
  {\begin{array}{l}\sem{\coctx{\Gamma}{\cclrd{r}} \vdash \lambda y. e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2}\\[-0.25em]~\end{array}}
  {f}
  {\hspace{-0.5em}\begin{array}{l}
  \lambda ((x_1, \ldots, x_n), g_1) . \lambda (y, g_2) .\\[-0.25em]
  \quad f~((x_1, \ldots, x_n, y), g_1 \uplus g_2)
  \end{array}\hspace{-1em}~}
& (\emph{abs})
\\[1.5em]
\hspace{-0.8em}\emdefff
  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}
  {\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1}
  {\begin{array}{l}\sem{\coctx{\Gamma}{\cclrd{r}\cup\cclrd{s}\cup\cclrd{t}} \vdash e_1~e_2 : \tau_2}\\[-0.25em]~\end{array}}
  {f_1}
  {f_2}
  {\hspace{-0.5em}\begin{array}{l}
  \lambda (x, g) . \\[-0.25em]
  \quad(f_1~(x, \restr{g}{\cclrd{r}}))~(f_2~(x, \restr{g}{\cclrd{s}} ), \restr{g}{\cclrd{t}})
  \end{array}\hspace{-2em}~}
& (\emph{app})
\\[1.5em]
\hspace{-3.3em}\emdefff
  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \ident{num}}
  {\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_2}
  {\begin{array}{l}\sem{\coctx{\Gamma}{\cclrd{r \cup (s \setminus \{ \ident{?p} \})}} \vdash\\[-0.25em]
    \qquad\kvd{letdyn}~\ident{?p}=e_1~\kvd{in}~e_2 : \tau_2}\end{array}}
  {f_1}
  {f_2}
  {\hspace{-0.5em}\begin{array}{l}
  \lambda (x, g) . f_2~(x, \restr{g}{\cclrd{s} \setminus \cclrd{\{ \ident{?p} \}}} \,\uplus \\[-0.25em]
    \qquad \{ \ident{?p} \mapsto (f_1~(x, \restr{g}{\cclrd{r}})) \} )
  \end{array}\hspace{-1em}~}
& (\emph{letdyn})
\end{array}
\end{equation*}

\vspace{1em}
Assuming the following auxiliary definitions:
\begin{equation*}
\begin{array}{rcl}
 \restr{f}{\cclrd{r}} &=& \{ (p,v) \;|\; (p,v) \in f, \;p \in \cclrd{r} \}\\[-0.25em]
 f \uplus g &=& \restr{f}{\, \textit{dom}(f) \setminus \textit{dom}(g)} \cup g
\end{array}
\end{equation*}

\figcaption{Semantics of a language with implicit parameters}
\label{fig:applications-flat-implsem}
\vspace{-0.5em}
\end{figure}

@@

\paragraph{Semantics.}
Implicit parameters can be implemented by passing around a hidden dictionary that provides values
to the implicit parameters. Accessing a parameter then becomes a lookup in the dictionary and
the new \kvd{letdyn} construct extends the dictionary. To elucidate how such hidden dictionaries
are propagated through the program when using lambda abstractions and applications, we present a
simple semantics for implicit parameters. The goal here is not to prove properties of the language,
but simply to provide a better explanation. A detailed semantics in terms of indexed comonads is
shown in Chapter~\ref{ch:semantics}.

Given an expression $e$ of type $\tau$ that requires free variables $\Gamma$ and implicit parameters
$\cclrd{r}$, the semantics is a function that takes a product of variables from $\Gamma$ together
with a dictionary of implicit parameters and returns $\tau$:
%
\begin{equation*}
\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \cclrd{r} } \vdash e : \tau}
  ~:~ (\tau_1 \times \ldots \times \tau_n) \times (\cclrd{r} \rightarrow \ident{num}) \rightarrow \tau
\end{equation*}
%
The dictionary is represented as a function from $\cclrd{r}$ to $\ident{num}$. This means that it
provides a $\ident{num}$ value for all implicit parameters that are required according to the typing.
Note that the domain of the function is not the set of all possible implicit parameter names, but
only the finite subset of names that are required according to the typing.
